Unveiling Robustness Threats by Mining the Exception Stack Traces of over 600 Android Projects

### 2014-09-03 android-stacks.tex
all over

\title{Discovering/Unveiling Robustness Threats by Mining the Exception Stack Traces of over 600 Android Projects}

@@

\begin{abstract}
The number of mobile apps is increasing in a daily rate. If on the one hand they
extend phones capabilities far beyond of the basic calls, on the other hand they have to 
cope with an increasing number of exceptinal conditions 
(e.g., faults in underlying middleware or hardware; memory and battery restrictions). 
Therefore, mechanisms for exception detection and handling are not an optional add-on but a 
fundamental part of such apps. However, studies have shown that such mechanisms are the less 
understood and tested parts of the system. As a consequence they may affect the system the 
other way around, making the system even less robust, leading to failures such as uncaught exceptions 
and unintended handler actions (i.e., exceptions being mistakenly caught by existing handlers).
This paper reports a study which mined the stack traces embedded on 
159,048 issues reported on Android projects hosted in GitHub (482 projs.) 
and Google Code (157 projs.). 
The goal of this study was to: investigate the whether the exception stack traces
 can reveal common characteristics on the exception-related code that
may be favoring the introduction of the aforementioned failures, which would be responsible for 
decreasing the application robustness. We call such characteristics ``fragilities" of the exception-related code.
Overall 6,005 exception stack traces were extracted, and analyzed in 
combination with source code and bytecode analysis. 
Some fragilities consistently detected in this study include: 
(1) undocumented runtime exceptions thrown by third party code;
(2) undocumented checked exceptions thrown by Android native code;
(3) unexpected exception wrappings (e.g., Errors being wrapped in checked exceptions),
revealing that Java hybrid exception model is not fully used according to its purpose.
Moreover, about 50\% of all analyzed projects reported at least one NullPointerExeption
as the root cause of an exception stack trace.
This study provide important insights on how such fragilities impose challenges 
to application robustness not only in Android environment but also to other 
systems based on Java exception model.

@@

Therefore, techniques for error detection and handling are not  an optional add-on but a 
fundamental part of such apps.
% as any other modern system~\cite{bruntink2006discovering}. 
The Android platform uses the exception handling mechanism embedded
 in Java to deal with such conditions. Although the exception handling mechanism
~\cite{goodenough1975exception} is one of the most used schemes for
detecting and recovering from such exceptional conditions, studies (e.g.~\cite{miller1997issues,Robil00,shah2010understanding, garcia2007extracting,garcia2001comparative,cabral2007exception,coelho2011unveiling}),
have shown that it is generally poorly understood and the less tested parts of the system ~\cite{coelho2011unveiling}.
As a consequence they may affect the system the other way around.
The exception-related code may favor the introduction of failures such as uncaught exceptions ~\cite{jo2004uncaught, zhang2012amplifying}
and unintended handler actions ~/cite{miller1997issues} (i.e., exceptions being mistakenly caught by existing handlers) 
which make the system even less robust.

@@

Until recent years, the availability of data reporting such failures was scarce. It comprised only few large scale projects (e.g. the Mozilla
crash report dataset\footnote{https://code.google.com/p/promisedata/issues/detail?id=88}).
Recently we have observed a rapid increase in the number of open
source projects that make their version control systems as well as the issue tracking systems
publicly available in repository hosting sites such as GitHub and Google Code.

@@

This paper reports a study which perform a post mortem analysis of the exception stack traces embedded on issues
reported on Android projects hosted in GitHub and Google Code. 
The goal of this study was to: investigate the whether the reported exception stack traces
 can reveal common fragilities on the exception-related code which may be favoring 
the introduction of failures such as uncaught exceptions - and as a consequence
 contributing to deacrease the robustness of apps.
To guide this investigation we compiled general guidelines on how to use
exceptions proposed by Gosling~\cite{gosling2000java},
Wirfs-Brock~\cite{wirfs2006toward} and Bloch~\cite{bloch2008effective}.
Then, using a custom tool called ExceptionMiner,
 which we develop specifically for this study, we mine stack traces from the issues reported 
on 482 Android projects hosted in GitHub and 157 projects hosted in Google Code.
Overall 159,048 issues were analyzed and 6,005 stack traces were extracted from them.
The exception stack trace analysis was augmented with additional information
extracted from JDK, Android Plaform and the applications using byte code 
and source code analysis. Manual inspection was used to leverage the understanding 
of exception stack traces and support further discussions and insights.

@@

Such fragilities represent obstacles to the development of robust java-based applications as they make difficult and 
sometimes almost impossible to protect the code against the Uncaught Exceptions that lead to system crashes.
Although the study focused on Android open-source applications, these findings are not restricted to this context, since 
the above mentioned fragilities may happen in any Java-based system. 
For instance: the unexpected wrappings were performed by Java libraries and elements (e.g. reflection library and 
the static blocks); and the undocumented runtime exceptions were also found in Java libraries besides the Android platform.
The study results are relevant to: (i) Android and Java developers to familiarize common characteristics and fragilities of exception-related code 
which is the first step to help developers preventing and dealing with them; (ii) designers of languages and tools to consider ways of reducing 
such fragilities in Android and Java environments.

Hence, the main contributions of this study are as follows:
\begin{itemize}

  \item  It performs a large scale post morten analysis of exception stack traces reported on Android projects hosted in GitHub and Google Code.

  \item  It introduces ExceptionMiner, a tool developed to support cross-repository stack trace
    analysis.

@@

The remaining of this paper is organized as follows. Section 2 presents a
background on Android Platform and Java exception model. 
Section 3 presents the study design. Section 4 reports study findings. 
Section 5 provides further discussions and insights.
Section 6 presents the threats to validity associated to this study. Finally Section
7 describes the related work, and Section 8 concludes the paper and outlines
directions for future work.

@@

\subsection{The Android Platform} \label{sec:extypes}

Android is an open source platform for mobile devices based on Linux kernel,
which also contains  a core Java library, a set of libraries third-party libraries (e.g.  SQLite),
the Dalvik runtime environment, and the Application Framework (which provides 
higher-level services to the applications running on the platform) ~\cite{andguide}.
Android platform is mostly implemented in Java and uses Java Native Invocation (JNI) to access native C/C++ code. 
And as mentioned before, it reuses the embedded Java Exception model to detect and handle 
abnormal computation states.

The Android Application Framework supports an event-driven development model. 
Among the basic elements that comprise an Android application, we can cite the
Activity. Each application usually defines one Activity per application window.
 Each Activity defines a set of hook methods  (e.g., OnCreate(), onPause()) that are dispatched
automatically by the Android framework (by inversion of
control), in response to infrastructure and application events.

@@

In Java, an exception is represented according to the class hierarchy shown in
Figure~\ref{fig:exchier}.  According to it every exception is an
instance of the Throwable class, and can be of three kinds: the checked exceptions
(extends Exception), the runtime exceptions (extends RuntimeException) and errors
(extends Error)~\cite{gosling2000java}.

The checked exceptions must be specified on the method signature that may throw it, 
and the compiler statically checks if appropriate handlers are provided within the system.
Both runtime exceptions and errors are known as unchecked exceptions, because 
they do not need to be specified on method signatures and do not trigger any 
compile time checking.

@@

Some concerns are known as sources of faults in mobile development are: concurrency ~\cite{ama2012} backward compatibility  
~\cite{McDon13}, security  ~\cite{enck2011study,was2010} and resource management (IO, Memory, Batery) ~\cite{Zhang12}.
Moreover, the programming errors are also as a source of faults, as illustrated on the previous section.
However, there are exceptions that cannot be directly mapped to one of such concerns: either because they are too general 
(e.g., java.lang.Exception) and its name does not hold a reason for the exception to be thrown; or because they are related to other concerns 
 (e.g., specific to a given library).

@@

Hence, based on the inspection of the source code and Javadoc related to each exception,  and Java specification, we identified the concern
releated to each root exception. Than we could calculate the prevalence of a concern on the reported stack traces using the following metric.
Given that: Stacks is the complete set of stacks found in this study, Lc is the list of root causes related to a given concern (as identified on the study),
 and Stacks[Lc] the list of issues that mentions the concern. The occurence of a concern could be given by the formula: Stacks[Lc]/Stacks.

@@

This class is responsible for managing the life cycle of  Android Activities - 
invoking the life cycle methods (i.e. onCreate(), onPause()) for every Activity 
(window) of an Android app.
Figure ~\ref{fig:snippets} (a) presents a code snippet of one the methods of 
ActivityThrea,  the one responsible for invoking the Activities onCreate method. 
As we can see from lines 4-8,
 in case any exception is thrown in the context of such try-catch block, 
it is wrapped in a general Runtime exception and re-thrown.

@@

However, since the Activity life clycle method (i.e., callActivityOnCreate(...)) that is invoked in this block,
 do not include checked exceptions on its signatures. Hence they could not throw a checked exception to be handled 
in the method illustrated above. However the code inspection performed on such wrappings revelaed an
``undocumented checked exception" signaled by this life cycle method.

\emph{\textbf{Undocumented Checked Exceptions.}} There are usually
2 possible possibilities for a checked exception to be thrown by a method:
it can either be specified on the method signature or it can be wrapped in a 
runtime exception and re-thrown. However by inspectioning associated to the reported wrappings 
revealed the a third option: a checked exception thrown by a native method, which was not declared
on the method signature of the methods on the call chain composing the stack.

The code snippet in Figure ~\ref{fig:snippets} (d) illustrate this scenario.
 The java/lang/Class declares a native method named getDeclaredMethods. 
The Java-side declaration of of this method does not have any throws clause, 
leading programmers and the compiler to think that no checked exceptions can be thrown.
 However, the C-code implementation did throw a "checked exception" NoSuchMethodException, 
violating the declaration. The Java compiler, could not detect this violation because it does 
not perform static exception checking on native methods. This type of bug is hard to diagnose
because the developer usualy do not have access to the native implementations. 
Consequently, since it is not expected by the programmer, when such method throws 
this exception, such undocumented exception may remain
uncaught and cause the app to crash, or may be mistakenly handled by subsumption - 
The exception stack traces reporting this scenario is actually a real bug of Android 
Gingerbread version (which still accounts for 13.6\% of devices running Android).

@@

In this paper we present an exploratory study in which we mined the stack 
traces embedded in all issues defined in 482 Android projects hosted in Github and 
157 projects hosted in Google Code.

In this study, the information extracted from stack traces was used in combination 
with source code and bytecode analysis to pin-point fragilities on the exception-related
code. 
This study  consistently identifies common characteristics of reported 
exception stacks traces - e.g., the NullPointerException was the reported cause 
on 30,08\% of all analyzed exception stack traces, and was mentioned in approximately 
50\% of all projects analyzed. Moreover,  the cross-project analysis of exception stack traces 
enabled to consistently reveal  fragilities of the exception-related code: 
(i) undocumented runtime exceptions signaled by third party code,
(ii) undocumented checked exceptions thrown by Android native code
(iii) unexpected exception wrappings (e.g., Errors being wrapped in checked exceptions) 
revealing that Java hybrid exception model is not fully used according to its purpose.
Such fragilities detected in our study can negatively affect the 
robustness of not only Android application but also Java-based 
which can deal with the same fragilities pointed in thid work. 
Our results calls the attention for the need of tool and language support to help 
developers when dealing with such problems.

>>>>

\title{Unveiling Robustness Threats by Mining the Exception Stack Traces of over 600 Android Projects}

@@

\begin{abstract}
The number of mobile apps is increasing in a daily rate. If on the one hand they
extend phones capabilities far beyond of the basic calls, on the other hand they have to 
cope with an increasing number of exceptinal conditions 
(e.g., faults in underlying middleware or hardware; memory and battery restrictions). 
Therefore, mechanisms for exception detection and handling are not an optional add-on but a 
fundamental part of such apps. However, studies have shown that such mechanisms are the less 
understood and tested parts of the system. As a consequence they may affect the system the 
other way around, making the system even less robust, leading to failures such as uncaught exceptions 
and unintended handling (i.e., exceptions being mistakenly caught by existing handlers).
This paper reports a study which mined the stack traces embedded on 
159,048 issues reported on Android projects hosted in GitHub (482 projs.) 
and Google Code (157 projs.). 
The goal of this study was to: investigate the whether the exception stack traces
 can reveal common characteristics on the exception-related code that
may be favoring the introduction of the aforementioned failures, which would be responsible for 
decreasing the application robustness. We call such characteristics ``fragilities" of the exception-related code.
Overall 6,005 exception stack traces were extracted, and analyzed in 
combination with source code and bytecode analysis. 
Some fragilities consistently detected in this study include: 
(1) undocumented runtime exceptions thrown by third party code;
(2) undocumented checked exceptions thrown by JNI native code;
(3) unexpected exception wrappings (e.g., Errors being wrapped in checked exceptions),
revealing that Java hybrid exception model is not fully used according to its purpose.
Moreover, about 50\% of all analyzed projects reported at least one NullPointerExeption
as the root cause of an exception stack trace.
This study provide important insights on how such fragilities impose challenges 
to application robustness not only in Android environment but also to other 
systems based on Java exception model.

@@

Therefore, techniques for error detection and handling are not  an optional add-on but a 
fundamental part of such apps.
% as any other modern system~\cite{bruntink2006discovering}. 
The Android platform uses the exception handling mechanism embedded
 in Java to deal with such conditions. Although the exception handling mechanism
~\cite{goodenough1975exception} is one of the most used schemes for
detecting and recovering from such exceptional conditions, studies (e.g.~\cite{miller1997issues,Robil00,shah2010understanding, garcia2007extracting,garcia2001comparative,cabral2007exception,coelho2011unveiling}),
have shown that it is generally poorly understood and the less tested parts of the system ~\cite{coelho2011unveiling}.
As a consequence they may affect the system the other way around.
The exception-related code may favor the introduction of failures such as uncaught exceptions ~\cite{jo2004uncaught, zhang2012amplifying}
and unintended handling ~/cite{miller1997issues} (i.e., exceptions being mistakenly caught by existing handlers) 
which make the system even less robust.

@@

Until recent years, the availability of data, reporting such failures, was scarce. 
It comprised only few large scale projects (e.g. the Mozilla crash report dataset\footnote{https://code.google.com/p/promisedata/issues/detail?id=88}).
Recently we have observed a rapid increase in the number of open
source projects that make their version control systems as well as the issue tracking systems
publicly available in repository hosting sites such as GitHub and Google Code.

@@

This study performs a post mortem analysis of the exception stack traces embedded on issues
reported on Android projects hosted in GitHub and Google Code. 
The goal of this study was to: investigate the whether the reported exception stack traces
 can reveal common fragilities on the exception-related code which may be favoring 
the introduction of failures such as uncaught exceptions - and as a consequence
 contributing to deacrease the robustness of apps.
To guide this investigation we compiled general guidelines on how to use
exceptions proposed by Gosling~\cite{gosling2000java},
Wirfs-Brock~\cite{wirfs2006toward} and Bloch~\cite{bloch2008effective}.
Then, using a custom tool called ExceptionMiner,
 which we develop specifically for this study, we mine stack traces from the issues reported 
on 482 Android projects hosted in GitHub and 157 projects hosted in Google Code.
Overall 159,048 issues were analyzed and 6,005 stack traces were extracted from them.
The exception stack trace analysis was augmented with additional information
extracted from JDK, Android Plaform and the applications using byte code 
and source code analysis. Manual inspection was used to leverage the understanding 
of exception stack traces and support further discussions and insights.

@@

As we can see, although the study was based on a subset of Android projects, 
the findings are not limited to this setting. The detected fragilities make difficult
or sometimes infeasible for a Java developer to prevent failures such as 
uncaught exceptions (e.g., thrown silently by third-party utility code), and 
unintended handling ( e.g., when an Error is wrapped in a checked Exception).
Thereore, they impose obstacles not only to the developement of robust Android apps, 
but also to the developement of any robust Java-based system. 
Hence, the study results are relevant to: (i) Android and 
Java developers to familiarize common characteristics and fragilities of exception-related code 
which is the first step to help developers preventing and dealing with them; and (ii) designers of languages and tools 
to consider ways of reducing such fragilities in Android and Java environments.
The main contributions of this study is twofold:
\begin{itemize}

  \item  It performs a large scale post morten analysis of exception stack traces reported on Android projects hosted in GitHub and Google Code.
  \item  It introduces ExceptionMiner, a tool developed to support exception stack trace mining and analysis.
    analysis.

@@

The remaining of this paper is organized as follows. Section 2 presents a
background on Android Platform and Java exception model. 
Section 3 presents the study design. Section 4 reports the study findings. 
Section 5 provides further discussions and insights.
Section 6 presents the threats to validity associated to this study. Finally Section
7 describes the related work, and Section 8 concludes the paper and outlines
directions for future work.

@@

\subsection{The Android Platform} \label{sec:extypes}
Android is an open source platform for mobile devices based on Linux kernel,
which also comprises a subset of the core Java library, a set of third-party libraries (e.g.  SQLite),
the Dalvik runtime environment, and the Application Framework (which provides 
higher-level services to the applications running on the platform) ~\cite{andguide}.

%The Android platform is mostly implemented in Java and uses Java Native Invocation (JNI) to access native C/C++ code. 
%And as mentioned before, it reuses the embedded Java Exception model to detect and handle 
%abnormal computation states.

The Android Application Framework supports an event-driven development model. 
Among the basic elements that comprise an Android application, we can cite the
 Activity class. The Activity is the main class of an Android app (there is usualy one
Activity per application window). Each Activity defines a set of hook methods  
(e.g., OnCreate(), onPause()) that are dispatched automatically by the Android 
framework (by inversion of control), in response to infrastructure and application events.

@@

In Java, an exception is represented according to the class hierarchy shown in
Figure~\ref{fig:exchier}.  According to it every exception is an
instance of the Throwable class, and can be of three kinds: the checked exceptions
(extends Exception), the runtime exceptions (extends RuntimeException) and errors
(extends Error)~\cite{gosling2000java}. The checked exception received such name
 because they must be declared on the method's exception interface (i.e., a list of exceptions that a method 
might raise during its execution) and the compiler statically checks if
 appropriate handlers are provided within the system.
Both runtime exceptions and errors are also known as unchecked exceptions, because 
they do not need to be specified on method exception interface and do not trigger any 
compile time checking.

@@

Some concerns are known as sources of faults in mobile development are: concurrency ~\cite{ama2012} backward compatibility  
~\cite{McDon13}, security  ~\cite{enck2011study,was2010} and resource management (IO, Memory, Batery) ~\cite{Zhang12}.

%Moreover, the programming errors are also as a source of faults, as illustrated on the previous section.

However, there are exceptions that cannot be directly mapped to one of such concerns. Either because they are
related to programming mistakes - as illustrated on the previously. Or because they are too general (i.e.,  java.lang.Exception,
java.lang.RuntimeException, java.lang.Error). Or also because they are related to other concerns (e.g., specific to a given library).

@@

Hence, based on the inspection of the source code and Javadoc related to each exception,  and Java specification, we identified the concern
releated to each root exception. Than we could calculate the prevalence of a concern on the reported stack traces using the following metric.
Given that: Stacks is the complete set of stacks found in this study, Lc is the list of root causes related to a given concern (as identified on the study),
 and Stacks[Lc] the list of issues that mentions the concern. The prevalence of a concern could be given by the formula: Stacks[Lc]/Stacks.

@@

This class is responsible for managing the life cycle of  Android Activities - 
invoking the life cycle methods (i.e. onCreate(), onPause()) for every Activity 
(window) of an Android app.
Figure ~\ref{fig:snippets} (a) presents a code snippet of one the methods of 
ActivityThrea,  the one responsible for invoking the Activitiy's onCreate method. 
As we can see from lines 4-8,
 in case any exception is thrown in the context of such try-catch block, 
it is wrapped in a general Runtime exception and re-thrown.

@@

%However, since the Activity life clycle method (i.e., callActivityOnCreate(...)) that is invoked in this block,
% do not include checked exceptions on its signatures. Hence they could not throw a checked exception to be handled 
%in the method illustrated above. However the code inspection performed on such wrappings revelaed an
%``undocumented checked exception" signaled by this life cycle method.

%\emph{\textbf{Undocumented Checked Exceptions.}} There are usually
%2 possible for a checked exception to be thrown by a method:
%it can either be specified on the method signature or it can be wrapped in a 
%runtime exception and re-thrown.

Inspectioning the source code, of the methods involved in this kind of wrapping,
revealed another fragility: a checked exception thrown by a native method and not
declared on the exception interface of thes methods signaling them. Such native method 
was defined on Android Platform -  which uses Java Native Invocation (JNI) to access 
native C/C++ code.

The code snippet in Figure ~\ref{fig:snippets} (d) illustrates this fragility.
 The java/lang/Class declares a native method named getDeclaredMethods. 
The Java-side declaration of of this method does not have any throws clause, 
leading programmers and the compiler to think that no checked exceptions can be thrown.
 However, the C-code implementation did throw a "checked exception" NoSuchMethodException, 
violating the declaration. The Java compiler, could not detect this violation because it does 
not perform static exception checking on native methods. This type of bug is hard to diagnose
because the developer usualy do not have access to the native implementations. 
Consequently, since it is not expected by the programmer, when such method throws 
this exception, such undocumented exception may remain
uncaught and cause the app to crash, or may be mistakenly handled by subsumption - 
The exception stack traces reporting this scenario is actually a real bug of Android 
Gingerbread version (which still accounts for 13.6\% of devices running Android).

@@

In this paper we present an exploratory study in which we mined the stack 
traces embedded in all issues defined in 482 Android projects hosted in Github and 
157 projects hosted in Google Code. Overall it included 6,005 exception stack traces.
In this study, the information extracted from stack traces was used in combination 
with source code and bytecode analysis to identify common characteristics of exception 
stack traces and pin-point fragilities on the exception-related code. 
We call ``fragilities" characteristics on the exception-related code that favor the introduction
of  uncaught exceptions and unintended handling - which can contribute to 
degrade the robustness of applications.
Some fragilities consistently detected in this study were: 
(i) an abundance of NullPointerExceptions - it was mentioned as the root exception in approximately 
50\% of the analyzed projects. 
(ii) undocumented runtime exceptions thrown by third party code,
(iii) undocumented checked exceptions thrown by JNI native code,
(iv) unexpected exception wrappings (e.g., Errors being wrapped in checked exceptions) 
revealing that Java hybrid exception model is not fully used according to its purpose.
As we can see, such fragilities can negatively affect not only the robustness of Android application 
but also the robustness of any Java-based application. 
Our results calls the attention for the need of tool and language support to help 
developers when dealing with such fragilities.
